Postural changes mediate children's visual access to social information

### 2018-01-16 writeup-cogsci-jan14/cogsci2018_faceOnly.Rmd
more edits

abstract: 
  "The ability to process social cues--including eye gaze--is a critical component of children's early language and cognitive development. However, as children reach their first birthday, they begin to locomote themselves, walking and exploring their visual environment in an entirely new way. How do these postural and locomotive changes affect children's access to the social information relevant for word-learning? Here, we explore this question by using head-mounted cameras to record infants' (8-16 months of age) egocentric visual perspective and modern computer vision algorithms to detect the proportion of faces in infants' environments.  We find that infant's postural changes largely mediate infants' access to faces, suggesting that these postural and locomotive developments facilitate infants' emerging linguistic and social capacities. Broadly, we suggest that the combined use of head-mounted cameras and the application of novel deep learning algorithms is a promising avenue for understanding the statistics of infants' visual and linguistic experience. "

@@

```{r, libraries}
library(knitr)
knitr::opts_chunk$set(fig.width=6, fig.height=5, fig.crop = FALSE, 
                      fig.path='figs/', echo=FALSE, warning = FALSE, 
                      cache=TRUE, message = FALSE, sanitize = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(xtable)
library(lubridate)
library(langcog)
library(ggplot2)
library(lme4)
library(gridExtra)
```
# Introduction
Children are remarkably skilled language learners, connecting arbitrary labels (“cup”) with specific visual concepts at a rapid pace through the first two years of life. However, children do not learn words in a vacuum but in a rich social environment, where social cues provided by speakers (e.g., eye-gaze) provide strong scaffolding for this learning process. Indeed, children’s ability to effectively process these social cues may be a key factor in their early language development. For example,in a longitudinal study, children’s level of joint engagement with their mother was found to predict both their receptive and productive vocabularies [@carpenter1998].  More recently, 10 month-olds who follow an adult’s gaze in an experimental context have larger vocabularies at 18 months [@brooks2005] and throughout the second year of life ]@brooks2008].

However, as children are learning their first words, their view of the world is also changing radically [@adolph2007]. Infant’s motor abilities improve dramatically near the end of the first year of life,  allowing them to locomote independently and to determine what they see. These motor changes have drastic consequences for what children see: crawling and walking infants have radically different views of the world. During spontaneous play in a laboratory playroom, toddlers are more likely to look at the floor while crawling than while walking [@franchak2011]; walking vs. crawling infants tend to have full visual access to their environment and the people in it [@kretch2014].

@@

Recent work has begun to use egocentric, head-mounted cameras to document the visual experiences of infants and children — which even for walking children are radically different than the adult perspective (and not easily predicted by our own adult intuitions; [@clerkin2017;@franchak2011, @yoshida2008] ). Children’s views tend to be restricted and to be dominated by objects and hands much more than that of adults [@yoshida2008], and both computational and empirical work suggest that this restricted viewpoint may be more effective for learning what objects and their labels than the comparable adult perspective [ @bambach2017;@yurovsky2012]. Further, recent work also suggests dramatic changes in the child’s perspective over the first two years of life, as views transition from primarily containing close up view of faces to capturing views of hands paired with the objects they are acting on [@fausey2016].

Here, we directly ask whether the postural changes that infants experience as they reach their first birthday change the availability of the social information relevant for word learning. To do so, we recorded the visual experience of a group of infants and children using head-mounted cameras during a brief laboratory free-play session; children’s posture and their orientation to their caregiver were also recorded from a third-person perspective and hand-annotated. We capitalize on recent improvements in face detection algorithms [@zhang2016, @cao2017realtime] and analyze the frequences of faces in the child’s visual environment.

@@

After the child was comfortable with wearing the headcam, the child and caregiver were shown to a playroom for the free-play session--the focus of the current study. Parents were shown a box containing three pairs of novel and familiar objects (e.g., a ball and a feather duster, named a "zem"), and were instructed to play with the object pairs with their child one at a time, "as they typically would." All parents confirmed that their child had not previosuly seen the novel toys and were instructed to use the novel labels to refer to the novel toys.

@@

### Posture and Orientation Annotation
We created a set of custom annotations that described the child's physical posture (e.g. standing) and the orientation of the caregiver relative to the child (e.g. far away). The child's posture was categorized as being held/carried, prone (crawling or lying), sitting, or standing. The caregiver's orientation was characterized as being close to the child, farther from the child, and a global category of caregiver behind the child. For the first two annotations (close/far from the child), the caregiver could either be to the the front or to the side of the child. All annotations were made using OpenSHAPA/Datavyu software  [@adolph2012], and times when the child was out of view of the tripod camera was marked as uncodable and was excluded from these annotations.

## Face Detection
Measuring the availability of caregiver’s faces across development was an important goal of the study. We thus used a total of two face detection systems to explore this hypothesis and to avoid the cost of hand-annotating every frame. We capitalized on recent improvements in computer vision, testing the performance of two algorithms that extract face information in distinct ways. Results for the evaluation of these three systems are shown in Table 2.

@@

The second algorithm was a CNN-based pose detector that provided the locations of 18 body parts (ears, nose, wrists, etc) (@cao2017realtime, @simon2017hand, @wei2016cpm) avaliable at https://github.com/CMU-Perceptual-Computing-Lab/openpose. The system uses a CNN for initial anatomical detection and subsequently applies part affinity fields (PAFs) for part association, producing a series of body part candidates. The candidates are then matched to a single individual and finally assembled into a pose. For the purposes of our investigation we only made use of the body parts relevant to the face (ears, eyes, nose). We operationalized face detections as any frames containing a nose, as any half of a face necessarily contains a nose.  
<!-- we operationalized hand detections as any frames containing any of the 21 points detected on either the left or the right hand, and we operationalized wrist detections any frames containing either the left or the right wrist.   -->

### Detector evaluation
In order to evaluate the performance of these detectors, we constructed a "gold set" of frames by hand labeling both a sample of frames with a high density of faces (as reported by the detectors) and a random sample from the remaining frames. This was done so as to not bias our evaluation by the relatively rare appearance of faces in the dataset. A face was present in a frame if at least half of the face was showing. Precision (hits / hits + false alarms), recall (hits / hits + misses), and F-score (harmonic mean of previous measures) were calculated and are reported in Table 2.
Both detectors performed relatively well on the gold set, with MTCNN outperforming OpenPose on the random sample but trailing behind in the high density sample. Due to high performance of these two detectors we have included the results of both in the analyses to follow.

@@

# Results
First, we report developmental shifts in infants' posture and their orientation to their caregiver. Then, we explore how these changes influence children's visual access to faces across this developmental time range. Finally, we explore how these changes impact the accessibility of faces during labeling events.

@@

```{r orientation, fig.env="figure", fig.pos = "h", fig.align = "center", fig.width=3, fig.height=2, fig.cap = "Proportion time that infants in each age group spent in each posture (left panel) and orientation relative to their caregiver (right  panel)." }

@@

```{r posture, fig.env="figure", fig.pos = "h", fig.align = "center", fig.width=3, fig.height=2, fig.cap = "Proportion time that infants in each age group spent in each posture (left panel) and orientation relative to their caregiver (right  panel)." }
ggplot(posture, aes(x = factor(age.grp), y = prop.time, col = posture)) + 
  geom_boxplot() + 
  ylab("Proportion Time") + 
  xlab("Age (months)") + 
  theme_bw()
```

@@

We first examined the proportion of face detections across age; a full summary can be seen in Figure \ref{fig:detsbyAge}. We observed  a slight U-shaped function both when analyzing the output of the MTCNN and OpenPose detectors, such that 12-month-olds appeared to experience slighly fewer faces faces than 8 or 16-month-olds.

@@

To formalize these observations, we fit a generalized logistic mixed-effect model to all with the presence/absence of a face on every frame as the dependent variable, and participant's age, orientation, and posture as predictors. Interactions between predictors were not included as this maximal model failed to converge. A summary of the coefficients of this model can be found in Table 3.  While age remained a significant predictor even when accounting for the effects of infants' posture and caregivers' orientation, it held relatively less predictive power. Overall, these results confirm that infant's access to faces is heavily infleunced by their own posture and their caregivers orientation towards them.

@@

\begin{table}[H]
\centering
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
  \hline
(Intercept) & -4.846 & 0.077 & -62.682 & 0.000 \\ 
  Age & 0.170 & 0.061 & 2.766 & 0.006 \\ 
  Posture-Prone & -0.513 & 0.036 & -14.045 & 0.000 \\ 
  Posture-Sit & 0.648 & 0.034 & 18.853 & 0.000 \\ 
  Posture-Stand & 1.005 & 0.035 & 28.402 & 0.000 \\ 
  Orient.-Close & 1.549 & 0.021 & 72.757 & 0.000 \\ 
  Orient.-Far & 2.258 & 0.022 & 101.138 & 0.000 \\ 
   \hline
\end{tabular}
\caption{Results from GLMM model prediting the presence/absence of a face across the entire dataset (MTCNN detectors).} 
\end{table}

## Access to Faces During Labeling Events
Finally, we analyzed how face detections changed during object labeling events as a function of infant's posture and orientation.  Specfically, we analyzed a four-second window around each labeling event (e.g., "Look at the [zem]!"); these labeling events were hand-annotated and automatically synchronized with the frame-by-frame face detections. We again found that infant's posture impacted the degree to which they saw their caregivers face during a labeling event; infants who were sitting or standing were more likely to have access to their caregiver's face.

@@

# General Discussion
We use a head-mounted camera to explore how children's postural and locomotive development directly impacts their access to social information relevant for word-learning, here operationalized as the presence of the faces of their caregiver. We found that children's posture and orientation towards their caregiver changed systematically across age, and that all three of these factors dramatically impacted the proportion of faces that were avaliable in the child's visual field. Thus, infants postural development is mediating factor that explains age-related changes in the proportion of faces avaliable in infants' visual field.

This work also deploys novel advancements in computer vision to the study of developmental psychology. The field of object detection and recognition has advanced dramatically in the past five years since the re-birth of deep learning algorithms @krizhevsky2012imagenet, creating a new generation of algorithmic tools. These tools are both substantially better equipped to deal with noisier, more complicated datasets and that can extract richer and more detailed information. Videos from the infant perspective provided sustantial challenges (e.g., partially occluded faces) for the classc models of face detedction (e.g., ViolaJones, @viola2004robust).  Further, as the headcam technologies employed here were inexpensive (~$60 a camera) and the computer vision algorithms freely avaliable, this method is a promising avenue for quanitfying the visual and social information avaliable to infant learners.

>>>>

abstract: 
  "The ability to process social cues--including eye gaze--is a critical component of children's early language and cognitive development. However, as children reach their first birthday, they begin to locomote themselves, walking and exploring their visual environment in an entirely new way. How do these postural and locomotive changes affect children's access to the social information relevant for word-learning? Here, we explore this question by using head-mounted cameras to record infants' (8-16 months of age) egocentric visual perspective and modern computer vision algorithms to detect the proportion of faces in infants' environments.  We find that infants' postural changes largely mediate infants' access to faces, suggesting that these postural and locomotive developments facilitate infants' emerging linguistic and social capacities. Broadly, we suggest that the combined use of head-mounted cameras and the application of novel deep learning algorithms is a promising avenue for understanding the statistics of infants' visual and linguistic experience."

@@

```{r, libraries}
library(knitr)
knitr::opts_chunk$set(fig.width=6, fig.height=5, fig.crop = FALSE, 
                      fig.path='figs/', echo=FALSE, warning = FALSE, 
                      cache=TRUE, message = FALSE, sanitize = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(xtable)
library(lubridate)
library(langcog)
library(ggplot2)
library(lme4)
library(gridExtra)
```
# Introduction
Children are remarkably skilled language learners, connecting arbitrary labels (“cup”) with specific visual concepts at a rapid pace through the first two years of life. However, children do not learn words in a vacuum but in a rich social environment, where social cues provided by speakers (e.g., eye-gaze) provide strong scaffolding for this learning process. Indeed, children’s ability to effectively process these social cues may be a key factor in their early language development. For example, in a longitudinal study, children’s level of joint engagement with their mother was found to predict both their receptive and productive vocabularies [@carpenter1998].  More recently, 10 month-olds who follow an adult’s gaze in an experimental context have larger vocabularies at 18 months [@brooks2005] and throughout the second year of life [@brooks2008].

However, as children are learning their first words, their view of the world is also changing radically [@adolph2007]. Infants' motor abilities improve dramatically near the end of the first year of life,  allowing them to locomote independently and to determine what they see. These motor changes have drastic consequences for what children see: crawling and walking infants have radically different views of the world. During spontaneous play in a laboratory playroom, toddlers are more likely to look at the floor while crawling than while walking [@franchak2011]; walking vs. crawling infants tend to have full visual access to their environment and the people in it [@kretch2014].

@@

Recent work has begun to use egocentric, head-mounted cameras to document the visual experiences of infants and children — which even for walking children are radically different from the adult perspective (and not easily predicted by our own adult intuitions [@clerkin2017;@franchak2011, @yoshida2008]). Children’s views tend to be restricted and to be dominated by objects and hands much more than that of adults [@yoshida2008], and both computational and empirical work suggest that this restricted viewpoint may be more effective for learning objects and their labels than the comparable adult perspective [ @bambach2017;@yurovsky2012]. Further, recent work also suggests dramatic changes in the child’s perspective over the first two years of life, as views transition from primarily containing a close up view of faces to capturing views of hands paired with the objects they are acting on [@fausey2016].

Here, we directly ask whether the postural changes that infants experience as they reach their first birthday change the availability of the social information relevant for word learning. To do so, we recorded the visual experience of a group of infants and children using head-mounted cameras during a brief laboratory free-play session; children’s posture and orientation relative to their caregiver were also recorded from a third-person perspective and hand-annotated. We capitalize on recent improvements in face detection algorithms [@zhang2016, @cao2017realtime] to analyze the frequences of faces in the child’s visual environment.

@@

After the child was comfortable wearing the headcam, the child and caregiver were shown to a playroom for the free-play session--the focus of the current study. Parents were shown a box containing three pairs of novel and familiar objects (e.g., a ball and a feather duster, named a "zem"), and were instructed to play with the object pairs with their child one at a time, "as they typically would." All parents confirmed that their child had not previously seen the novel toys and were instructed to use the novel labels to refer to the novel toys.

@@

### Posture and Orientation Annotation
We created a set of custom annotations that described the child's physical posture (e.g. standing) and the orientation of the caregiver relative to the child (e.g. far away). The child's posture was categorized as being held/carried, prone (crawling or lying), sitting, or standing. The caregiver's orientation was characterized as being close to the child, far from the child, and a global category of caregiver behind the child. For the first two annotations (close/far from the child), the caregiver could either be to the the front or to the side of the child. All annotations were made using OpenSHAPA/Datavyu software  [@adolph2012], and times when the child was out of view of the tripod camera were marked as uncodable and were excluded from these annotations.

## Face Detection
Measuring the availability of caregiver’s faces across development was an important goal of this study. We used a total of two face detection systems to explore infants' changing access to social information and to avoid the cost of hand-annotating every frame. We capitalized on recent improvements in computer vision algorithms, testing the performance of two state-of-the-art face detectors that used novel methods to extract face information.

@@

The second algorithm was a CNN-based pose detector that provided the locations of 18 body parts (ears, nose, wrists, etc.) (@cao2017realtime, @simon2017hand, @wei2016cpm) avaliable at https://github.com/CMU-Perceptual-Computing-Lab/openpose. The system uses a CNN for initial anatomical detection and subsequently applies part affinity fields (PAFs) for part association, producing a series of body part candidates. The candidates are then matched to a single individual and finally assembled into a pose. For the purposes of our investigation we only made use of the body parts relevant to the face (ears, eyes, nose). We operationalized face detections as any frames containing a nose, as any half of a face necessarily contains a nose.  
<!-- we operationalized hand detections as any frames containing any of the 21 points detected on either the left or the right hand, and we operationalized wrist detections any frames containing either the left or the right wrist.   -->

### Detector evaluation
In order to evaluate the performance of these detectors, we constructed a "gold set" of frames by hand labeling both a sample of frames with a high density of faces (as reported by the detectors) and a random sample from the remaining frames. This was done so as to not bias our evaluation by the relatively rare appearance of faces in the dataset. A face was present in a frame if at least half of the face was showing. Precision (hits / hits + false alarms), recall (hits / hits + misses), and F-score (harmonic mean of previous measures) were calculated and are reported in Table 2.
Both detectors performed relatively well on the gold set, with MTCNN outperforming OpenPose on the random sample but trailing behind in the high density sample. Due to the high performance of these two detectors we have included the results of both in the analyses to follow.

@@

# Results
First, we report developmental shifts in infants' posture and their orientation relative to their caregiver. Then, we explore how these changes influence children's visual access to faces across this developmental time range. Finally, we explore how these changes impact the accessibility of faces during labeling events.

@@

```{r orientation, fig.env="figure", fig.pos = "h", fig.align = "center", fig.width=3, fig.height=2, fig.cap = "Proportion time that infants in each age group spent in each orientation relative to their caregiver." }

@@

```{r posture, fig.env="figure", fig.pos = "h", fig.align = "center", fig.width=3, fig.height=2, fig.cap = "Proportion time that infants in each age group spent in each posture." }
ggplot(posture, aes(x = factor(age.grp), y = prop.time, col = posture)) + 
  geom_boxplot() + 
  ylab("Proportion Time") + 
  xlab("Age (months)") + 
  theme_bw()
```

@@

We first examined the proportion of face detections across age; a full summary can be seen in Figure \ref{fig:detByAge}. We observed  a slight U-shaped function both when analyzing the output of the MTCNN and OpenPose detectors, such that 12-month-olds appeared to experience slighly fewer faces faces than 8 or 16-month-olds.

@@

To formalize these observations, we fit a generalized logistic mixed-effect model to all with the presence/absence of a face on every frame as the dependent variable, and participant's age, orientation, and posture as predictors. Interactions between predictors were not included as this maximal model failed to converge. A summary of the coefficients of this model can be found in Table 3.  While age remained a significant predictor even when accounting for the effects of infants' posture and caregivers' orientation, it held relatively less predictive power. Overall, these results confirm that infants' access to faces is heavily infleunced by their own posture and their caregiver's orientation towards them.

@@

\begin{table}[H]
\centering
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & z value & Pr($>$$|$z$|$) \\ 
  \hline
(Intercept) & -4.846 & 0.077 & -62.682 & 0.000 \\ 
  Age & 0.170 & 0.061 & 2.766 & 0.006 \\ 
  Posture-Prone & -0.513 & 0.036 & -14.045 & 0.000 \\ 
  Posture-Sit & 0.648 & 0.034 & 18.853 & 0.000 \\ 
  Posture-Stand & 1.005 & 0.035 & 28.402 & 0.000 \\ 
  Orient.-Close & 1.549 & 0.021 & 72.757 & 0.000 \\ 
  Orient.-Far & 2.258 & 0.022 & 101.138 & 0.000 \\ 
   \hline
\end{tabular}
\caption{Results from GLMM model predicting the presence/absence of a face across the entire dataset (MTCNN detectors).} 
\end{table}

## Access to Faces During Labeling Events
Finally, we analyzed how face detections changed during object labeling events as a function of infant's posture and orientation.  Specfically, we analyzed a four-second window around each labeling event (e.g., "Look at the [zem]!"); these labeling events were hand-annotated and automatically synchronized with the frame-by-frame face detections. We again found that infant's posture impacted the degree to which they saw their caregiver's face during a labeling event; infants who were sitting or standing were more likely to have access to their caregiver's face.

@@

# General Discussion
We use a head-mounted camera to explore how children's postural and locomotive development directly impacts their access to social information relevant for word-learning, here operationalized as the presence of the faces of their caregiver. We found that children's posture and orientation towards their caregiver changed systematically across age, and that all three of these factors dramatically impacted the proportion of faces that were avaliable in the child's visual field. Thus, infants' postural development is a mediating factor that explains age-related changes in the proportion of faces avaliable in infants' visual field.

This work also deploys novel advancements in computer vision to the study of developmental psychology. The field of object detection and recognition has advanced dramatically in the past five years since the re-birth of deep learning algorithms @krizhevsky2012imagenet, creating a new generation of algorithmic tools. These tools are substantially better equipped to deal with noisier, more complicated datasets and can extract richer and more detailed information. Videos from the infant perspective provided substantial challenges (e.g., partially occluded faces) for the classic models of face detection (e.g., ViolaJones, @viola2004robust).  Further, as the headcam technologies employed here were inexpensive (~$60 a camera) and the computer vision algorithms freely avaliable, this method is a promising avenue for quanitfying the visual and social information avaliable to infant learners.
